---
title: "Weather Events and their impact on Economy and Public Health"
author: "John Doe"
date: "20th january 2015"
output:
  html_document:
    fig_width: 10
---

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. Specially tornadoes and floods have caused lots of unnecessary deaths and economic loses.

## Data Processing

**Performed steps**:  
* Get and load the data  
* Remove unused information

```{r loading and cleaning the data,cache=TRUE}

# check to see if the data set exists; if not, download it
if (!file.exists('repdata-data-StormData.csv.bz2')) {
    # download the zip file and unzip
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                  "repdata-data-StormData.csv.bz2", mode="wb")
}
# load the data from the bz2 file
storm_data_connection <- bzfile("repdata-data-StormData.csv.bz2", open = "r")
storm_data <- read.table(storm_data_connection, header = TRUE, sep = ",")
close(storm_data_connection)
remove(storm_data_connection)

# remove column information that we do not use to save space
good_columns <- c("EVTYPE",                  # Event type
                  "FATALITIES", "INJURIES",  # Fatalities and Injuries
                  "PROPDMG", "PROPDMGEXP",   # Property damange and exponential
                  "CROPDMG", "CROPDMGEXP")   # Crop damage and exponential
storm_data <- storm_data[,good_columns]
summary(storm_data)
```

* Add new properties for damage and crop damage to get the total damage
* Calculate fatalities and injuries (total amount)

```{r Calculate the damage, cache=TRUE}
#Add property damage and crop damage to get the total damage
levels(storm_data$PROPDMGEXP) <- c(
  "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", "1", 
  "1000000000", "1", "1", "1000", "1000000", "1000000")
levels(storm_data$CROPDMGEXP) <- c(
  "1", "1", "1", "1", "1000000000", "1000", 
  "1000", "1000000", "1000000")
storm_data$PROPDMG <- storm_data$PROPDMG * 
  as.integer(as.character(storm_data$PROPDMGEXP))
storm_data$CROPDMG <- storm_data$CROPDMG * 
  as.integer(as.character(storm_data$CROPDMGEXP))

# add up damage
storm_data$DAMAGE <- storm_data$PROPDMG + storm_data$CROPDMG
# damage in billions
storm_data$DAMAGEBIL <- sapply(storm_data$DAMAGE, function(x) x/1000000000)
# Calculate fatalities and injuries (total amount)
storm_data$HEALTH <- storm_data$INJURIES + storm_data$FATALITIES
# health in kilos
storm_data$HEALTHK <- sapply(storm_data$HEALTH, function(x) x/1000)
```

## Results
### Population health impacts
Question: Across the United States, which types of events are most harmful with respect to population health?

```{r Impact plot about population health issues because of weather events}
total <- sort(tapply(storm_data$HEALTHK, storm_data$EVTYPE, sum),decreasing = TRUE)
count <- 20;
par(pin=c(10,8)) # set figure size in inches
par(las=2) # make label text perpendicular to axis
par(mar=c(5,15,4,2)) # increase y-axis margin.
par(mgp=c(0.5, 2, 0)) # set margins
barplot(head(total,count),
        horiz=TRUE,
        main="Top 20 most unhealthy weather events",
        ylab="Event",
        xlab="Fatalities and injuries (thousands)",
        xlim=c(0,100),
        border="black",
        col=c("red"))
max(total)
```

Tornados caused much more heath damage than the others.

### Economy impacts
Question: Across the United States, which types of events have the greatest economic consequences?

```{r Plot about economy impacts}
total <- sort(tapply(storm_data$DAMAGEBIL, storm_data$EVTYPE, sum), decreasing = TRUE)
par(pin=c(10,8)) # set figure size in inches
par(las=2) # make label text perpendicular to axis
par(mar=c(5,15,4,2)) # increase y-axis margin.
par(mgp=c(0.5, 2, 0)) # set margins
barplot(head(total,count),
        horiz=TRUE,
        main="Top 20 most expensive weather events",
        ylab="Event",
        xlab="Billion USD",
        xlim=c(0,160),
        border="black",
        col=c("red"),
        pin=c(10,8))
max(total)
```

Floods caused much more costs than the other weather events.